Transferring Value Functions by Variational Methods

### 2018-05-17 NIPS 2018/main.tex
Minor changes to Section 3

Although the Gaussian assumption of the previous section is very appealing as it allows for a simple and efficient way of computing the variational objective and its gradients, we believe that such assumption almost never holds in practice. In fact, even for families of tasks in which the reward and transition models are Gaussian, the $Q$-values might be far from it. Depending on the family of tasks under consideration and, since we are learning a distribution over weights, on the chosen function approximator, the prior might have arbitrarily complex shapes. When the information loss due to the Gaussian approximation becomes too severe, the algorithm is likely too fail at capturing any similarities between the tasks. We now propose a variant to successfully solve this problem, while keeping the algorithm simple and efficient enough to be applied in practice.

Given the source tasks' weights $\mathcal{W}_s$, we model our estimated prior as a mixture of Gaussians with one equally weighted isotropic Gaussian centered at each weight: $p(\bm{w}) = \frac{1}{\abs{\mathcal{W}_s}}\sum_{\bm{w}_s \in \mathcal{W}_s}\mathcal{N}(\bm{w}|\bm{w}_s,\sigma_p^2\bm{I})$. This resembles a kernel density estimator \cite{scott2015multivariate} and, due to its nonparametric nature, it allows capturing arbitrarily complex distribution. Consistently with the prior, we model our approximate posterior as a mixture of Gaussians. However, we allow a different number of components (typically much less than the prior's) and we adopt full covariances instead of only diagonals, so that our posterior has the potential to match complex distributions with less components. Using $C$ components, our posterior is $q_{\bm{\xi}}(\bm{w}) = \frac{1}{C}\sum_{i=1}^C\mathcal{N}(\bm{w}|\bm{\mu}_i,\bm{\Sigma}_i)$, with variational parameters $\bm{\xi} = (\bm{\mu}_1,\dots,\bm{\mu}_C,\bm{\Sigma}_1,\dots,\bm{\Sigma}_C)$. Once again, we learn Cholesky factors instead of full covariances.

Although this new model has the potential to capture much more complex distributions, it poses a major complication: the KL divergence between two mixture of Gaussians has no closed-form expression. To solve this issue, we rely on an upper bound to such quantity, so that negative ELBO still upper bounds the KL between the approximate and true posterior. Among the many upper bounds available in the literature, we adopt the one proposed in \cite{hershey2007approximating}, which we report here for the sake of completeness. We refer the reader to the original paper for the proof.
\begin{theorem}[\cite{hershey2007approximating}]\label{th:ukl}
Let $p = \sum_{i} c_i^{(p)} f_i^{(p)}$ and $q = \sum_{j} c_j^{(q)} f_j^{(q)}$ be two mixture of Gaussian distributions, where $f_i^{(p)} = \mathcal{N}(\bm{\mu}_i^{(p)},\bm{\Sigma}_i^{(p)})$ denotes the i-th component of $p$, $c_i^{(p)}$ denotes its weight, and similarly for $q$. Introduce two vectors $\chi^{(1)}$ and $\chi^{(2)}$ such that $c_i^{(p)} = \sum_j \chi^{(2)}_{j,i}$ and $c_j^{(q)} = \sum_i \chi^{(1)}_{i,j}$. Then:
\begin{equation}
KL(p||q) \leq KL(\chi^{(2)}||\chi^{(1)}) + \sum_{i,j} \chi^{(2)}_{j,i}KL(f_i^{(p)}||f_j^{(q)})
\end{equation}
%\begin{equation}
%\chi^{(2)}_{i,j} = \frac{c_j^{(q)}\chi^{(1)}_{j,i}}{\sum_{i'} \chi^{(1)}%_{j,i'}},\ \ \ \ \chi^{(1)}_{j,i} = \frac{c_i^{(p)}\chi^{(2)}_{i,j}e^{-%KL(f_i^{(p)}||f_j^{(q)})}}{\sum_{j'}\chi^{(2)}_{i,j'}e^{-KL(f_i^{(p)}||f_{j'}%^{(q)})}}
%\end{equation}
\end{theorem}
Our new algorithm replaces the KL with the above-mentioned upper bound. Each time we require its value, we have to recompute the parameters $\chi^{(1)}$ and $\chi^{(2)}$ that tighten the bound. As shown in \cite{hershey2007approximating}, this can be achieved by a simple fixed-point procedure. Finally, both terms in the objective are now linear combinations of functions of the variational parameters of different components, and their gradients are easily derived from the ones of the Gaussian case. We report the derivation in App. \ref{}.

@@

From Sections \ref{sec:gvt} and \ref{sec:mgvt}, we know that differentiating the negative ELBO $\mathcal{L}$ w.r.t. $\bm{\xi}$ requires differentiating $\norm{B(\bm{w})}_D^2$ w.r.t. $\bm{w}$. Unfortunately, the TD error is well-known to be non-differentiable due to the presence of the max operator. This rarely represents a problem since typical value-based algorithms are actually semi-gradient methods, i.e., they do not differentiate the targets (see, e.g., Chapter 11 of \cite{sutton1998reinforcement}). However, our transfer settings are rather different than common RL. In fact, our algorithm is likely to start from $Q$-functions that are very close to an optimum, and the only thing that needs to be done is to adapt the weights in a direction of lower error (i.e., higher likelihood) so as to quickly converge to the solution of the task that is being solved. Unfortunately, this property cannot be guaranteed for most semi-gradient algorithms. Even worse, many online RL algorithms combined with complex function approximators (e.g., DQNs) are well-known to be unstable, especially when approaching an optimum, and require many tricks and tuning to work well \cite{schaul2015prioritized,van2016deep}. This is obviously an undesirable property in our case, as we only aim at adapting already good solutions. Thus, we consider using a residual gradient algorithm \cite{baird1995residual}. In order to differentiate the targets, we replace the optimal Bellman operator with the mellow Bellman operator introduced in \cite{asadi2017alternative}, which adopts a softened version of max called \textit{mellowmax}:
\begin{equation}
\mm_a Q_{\bm{w}}(s,a) = \frac{1}{\kappa} \log \frac{1}{|\mathcal{A}|}\sum_a e^{\kappa Q_{\bm{w}}(s,a)}
\end{equation}
where $\kappa$ is a hyperparameter and $|\mathcal{A}|$ is the number of actions. The mellow Bellman operator, which we denote as $\wt{T}$, has several appealing properties that make it suitable for our settings: (i) it converges to the maximum as $\kappa \rightarrow \infty$, (ii) it has a unique fixed point, and (iii) it is \textit{differentiable}. Denoting by $\wt{B}(\bm{w}) = \wt{T}Q_{\bm{w}} - Q_{\bm{w}}$ the Bellman residual w.r.t. the mellow Bellman operator $\wt{T}$, we have that the corresponding TD error, $\norm{\wt{B}(\bm{w})}_D^2$, is now differentiable with respect to $\bm{w}$. Although residual algorithms have guaranteed convergence, they are typically much slower than their semi-gradient counterpart. \cite{baird1995residual} proposed to project the gradient in a direction that achieves higher learning speed, while preserving convergence. This can be easily done by including a parameter $\psi\in[0,1]$ in the TD error gradient such that:
\begin{equation}
\nabla_{\bm{w}}\norm{\wt{B}(\bm{w})}_{D}^2 = \frac{2}{N}\sum_{i=1}^N b_i(\bm{w})\left( \gamma\psi\nabla_{\bm{w}}\mm_{a'}Q_{\bm{w}}(s_i',a') - \nabla_{\bm{w}}Q_{\bm{w}}(s_i,a_i)) \right)
\end{equation}
where $b_i(\bm{w}) = r_i + \gamma\mm_{a'}Q_{\bm{w}}(s_i',a') - Q_{\bm{w}}(s_i,a_i)$. Notice that $\psi$ trades-off between the semi-gradient ($\psi = 0$) and the full residual gradient ($\psi = 1$). A good criterion for choosing such parameter is to start with values close to zero (to have faster learning) and move to higher values when approaching the optimum (to guarantee converge).

>>>>

Although the Gaussian assumption of the previous section is very appealing as it allows for a simple and efficient way of computing the variational objective and its gradients, in practice it rarely allows us to describe the prior distribution accurately. In fact, even for families of tasks in which the reward and transition models are Gaussian, the $Q$-values might be far from being normally distributed. Depending on the family of tasks under consideration and, since we are learning a distribution over weights, on the chosen function approximator, the prior might have arbitrarily complex shapes. When the information loss due to the Gaussian approximation becomes too severe, the algorithm is likely to fail at capturing any similarities between the tasks. We now propose a variant to successfully solve this problem, while keeping the algorithm efficient and simple enough to be applied in practice.

Given the source tasks' weights $\mathcal{W}_s$, we model our estimated prior as a mixture of Gaussians with one equally weighted isotropic Gaussian centered at each weight: $p(\bm{w}) = \frac{1}{\abs{\mathcal{W}_s}}\sum_{\bm{w}_s \in \mathcal{W}_s}\mathcal{N}(\bm{w}|\bm{w}_s,\sigma_p^2\bm{I})$. This model resembles a kernel density estimator \cite{scott2015multivariate} and, due to its nonparametric nature, it allows capturing arbitrarily complex distributions. Consistently with the prior, we model our approximate posterior as a mixture of Gaussians. However, we allow a different number of components \todo{questo dettaglio si potrebbe spostare nella parte sperimentale o togliere del tutto} (typically much less than the prior's), and we adopt full covariances instead of only diagonals so that our posterior has the potential to match complex distributions with fewer components. Using $C$ components, our posterior is $q_{\bm{\xi}}(\bm{w}) = \frac{1}{C}\sum_{i=1}^C\mathcal{N}(\bm{w}|\bm{\mu}_i,\bm{\Sigma}_i)$, with variational parameters $\bm{\xi} = (\bm{\mu}_1,\dots,\bm{\mu}_C,\bm{\Sigma}_1,\dots,\bm{\Sigma}_C)$. Once again, we learn Cholesky factors instead of full covariances.

Although this new model has the potential to capture much more complicated distributions, it poses a significant complication: the KL divergence between two mixtures of Gaussians has no closed-form expression. To solve this issue, we rely on an upper bound to such quantity, so that negative ELBO still upper bounds the KL between the approximate and the exact posterior. Among the many upper bounds available in the literature, we adopt the one proposed in \cite{hershey2007approximating}, which we report here for the sake of completeness. We refer the reader to the original paper for the proof.
\begin{theorem}[\cite{hershey2007approximating}]\label{th:ukl}
Let $p = \sum_{i} c_i^{(p)} f_i^{(p)}$ and $q = \sum_{j} c_j^{(q)} f_j^{(q)}$ be two mixture of Gaussian distributions, where $f_i^{(p)} = \mathcal{N}(\bm{\mu}_i^{(p)},\bm{\Sigma}_i^{(p)})$ denotes the i-th component of $p$, $c_i^{(p)}$ denotes its weight, and similarly for $q$. Introduce two vectors $\chi^{(1)}$ and $\chi^{(2)}$ such that $c_i^{(p)} = \sum_j \chi^{(2)}_{j,i}$ and $c_j^{(q)} = \sum_i \chi^{(1)}_{i,j}$. Then:
\begin{equation}
KL(p||q) \leq KL(\chi^{(2)}||\chi^{(1)}) + \sum_{i,j} \chi^{(2)}_{j,i}KL(f_i^{(p)}||f_j^{(q)})
\end{equation}
%\begin{equation}
%\chi^{(2)}_{i,j} = \frac{c_j^{(q)}\chi^{(1)}_{j,i}}{\sum_{i'} \chi^{(1)}%_{j,i'}},\ \ \ \ \chi^{(1)}_{j,i} = \frac{c_i^{(p)}\chi^{(2)}_{i,j}e^{-%KL(f_i^{(p)}||f_j^{(q)})}}{\sum_{j'}\chi^{(2)}_{i,j'}e^{-KL(f_i^{(p)}||f_{j'}%^{(q)})}}
%\end{equation}
\end{theorem}
Our new algorithm replaces the KL with the above-mentioned upper bound. Each time we require its value, we have to recompute the parameters $\chi^{(1)}$ and $\chi^{(2)}$ that tighten the bound. As shown in \cite{hershey2007approximating}, we can use a simple fixed-point procedure for this purpose. Finally, both terms in the objective are now linear combinations of functions of the variational parameters of different components, and their gradients easily derive from the ones of the Gaussian case. We report the derivation in App. \ref{}.

@@

From Sections \ref{sec:gvt} and \ref{sec:mgvt}, we know that differentiating the negative ELBO $\mathcal{L}$ w.r.t. $\bm{\xi}$ requires differentiating $\norm{B(\bm{w})}_D^2$ w.r.t. $\bm{w}$. Unfortunately, the TD error is well-known to be non-differentiable due to the presence of the max operator. This issue rarely is a problem since typical value-based algorithms are semi-gradient methods, i.e., they do not differentiate the targets (see, e.g., Chapter 11 of \cite{sutton1998reinforcement}). However, our transfer settings are quite different from common RL. In fact, our algorithm is likely to start from $Q$-functions that are very close to an optimum and aims only to adapt the weights in some direction of lower error (i.e., higher likelihood) so as to quickly converge to the solution of the target task. Unfortunately, this property does not hold for most semi-gradient algorithms. Even worse, many online RL algorithms combined with complex function approximators (e.g., DQNs) are well-known to be unstable, especially when approaching an optimum, and require many tricks and tuning to work well \cite{schaul2015prioritized,van2016deep}. This property is clearly undesirable in our case, as we only aim at adapting already good solutions. Thus, we consider using a residual gradient algorithm \cite{baird1995residual}. To differentiate the targets, we replace the optimal Bellman operator with the mellow Bellman operator introduced in \cite{asadi2017alternative}, which adopts a softened version of max called \textit{mellowmax}:
\begin{equation}
\mm_a Q_{\bm{w}}(s,a) = \frac{1}{\kappa} \log \frac{1}{|\mathcal{A}|}\sum_a e^{\kappa Q_{\bm{w}}(s,a)}
\end{equation}
where $\kappa$ is a hyperparameter and $|\mathcal{A}|$ is the number of actions. The mellow Bellman operator, which we denote as $\wt{T}$, has several appealing properties that make it suitable for our settings: (i) it converges to the maximum as $\kappa \rightarrow \infty$, (ii) it has a unique fixed point, and (iii) it is \textit{differentiable}. Denoting by $\wt{B}(\bm{w}) = \wt{T}Q_{\bm{w}} - Q_{\bm{w}}$ the Bellman residual w.r.t. the mellow Bellman operator $\wt{T}$, we have that the corresponding TD error, $\norm{\wt{B}(\bm{w})}_D^2$, is now differentiable w.r.t. $\bm{w}$. Although residual algorithms have guaranteed convergence, they are typically much slower than their semi-gradient counterpart. \cite{baird1995residual} proposed to project the gradient in a direction that achieves higher learning speed, while preserving convergence. This projection is obtained by including a parameter $\psi\in[0,1]$ in the TD error gradient:
\begin{equation}
\nabla_{\bm{w}}\norm{\wt{B}_{\bm{w}}}_{D}^2 = \frac{2}{N}\sum_{i=1}^N b_i(\bm{w})\left( \gamma\psi\nabla_{\bm{w}}\mm_{a'}Q_{\bm{w}}(s_i',a') - \nabla_{\bm{w}}Q_{\bm{w}}(s_i,a_i)) \right),
\end{equation}
where $b_i(\bm{w}) = r_i + \gamma\mm_{a'}Q_{\bm{w}}(s_i',a') - Q_{\bm{w}}(s_i,a_i)$. Notice that $\psi$ trades-off between the semi-gradient ($\psi = 0$) and the full residual gradient ($\psi = 1$). A good criterion for choosing such parameter is to start with values close to zero (to have faster learning) and move to higher values when approaching the optimum (to guarantee convergence).
